---
title: "Looking_at_the_data_from_Moorea"
format: html
editor: visual
---

# Libraries
```{r}
library(tidyverse)
library(here)
library(dplyr)
```

Citation: 
Moorea Coral Reef LTER and A. Brooks. 2023. MCR LTER: Coral Reef: Long-term Population and Community Dynamics: Fishes, ongoing since 2005 ver 62. Environmental Data Initiative. https://doi.org/10.6073/pasta/75644add7e7f90c568bf5045264d359a (Accessed 2025-02-21).


#Load Data
Options: Abudefduf septemfasciatus <- dog fish tuna 104 observations
Acanthurus olivaceus <- orange band surgeon fish 379 obs
Naso lituratus <- orange spine unicornfish 1279 observations
Acanthurus triostegus <- manini 466
Carcharhinus melanopterus <- black tip reef shark 51 obs
```{r}
species_list <- read_csv(here('Data','MCR_LTER_Fish_Species_List_20230608.csv'))
annual_survey <- read_csv(here('Data','MCR_LTER_Annual_Fish_Survey_20230615.csv'))

fishies <- annual_survey |> filter(Taxonomy == "Naso lituratus")
```



```{r}
#just the orange spine unicorn fishies count per year
unicorn_count_avg <- fishies |> 
  group_by(Year) |>
  summarise(Average_Count = mean(Count, na.rm = TRUE))

unicorn_count_avg$Year <- as.numeric(unicorn_count_avg$Year)
ggplot(unicorn_count_avg, aes(x = Year, y = Average_Count)) +
  geom_line(color = "blue") + theme_classic()
```
Unicorn fish biomass per year
```{r}
unicorn_biomas_avg <- fishies |> 
  group_by(Year) |>
  summarise(Average_Biomass = mean(Biomass, na.rm = TRUE))

unicorn_biomas_avg$Year <- as.numeric(unicorn_biomas_avg$Year)

ggplot(unicorn_biomas_avg, aes(x = Year, y = Average_Biomass)) +
  geom_line(color = "blue") + theme_classic()

```

```{r}
#all fishies count
all_count_avg <- fishies |> 
  group_by(Year) |>
  summarise(Average_Count = mean(Count, na.rm = TRUE))

all_count_avg$Year <- as.numeric(all_count_avg$Year)

ggplot(all_count_avg, aes(x = Year, y = Average_Count)) +
  geom_line(color = "blue") + theme_classic()
```
Link to data: https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-mcr.6.62

```{r}
all_biomass_avg <- fishies |> 
  group_by(Year) |>
  summarise(Average_Biomass = mean(Biomass, na.rm = TRUE))

all_biomass_avg$Year <- as.numeric(all_biomass_avg$Year)

ggplot(all_biomass_avg, aes(x = Year, y = Average_Biomass)) +
  geom_line(color = "blue") + theme_classic()
```

