---
title: "Looking_at_the_data_from_Moorea"
format: html
editor: visual
---

# Libraries
```{r}
library(tidyverse)
library(here)
```

Citation: 
Moorea Coral Reef LTER and A. Brooks. 2023. MCR LTER: Coral Reef: Long-term Population and Community Dynamics: Fishes, ongoing since 2005 ver 62. Environmental Data Initiative. https://doi.org/10.6073/pasta/75644add7e7f90c568bf5045264d359a (Accessed 2025-02-21).


#Load Data
Options: Abudefduf septemfasciatus <- dog fish tuna 104 observations
Acanthurus olivaceus <- orange band surgeon fish 379 obs
Naso lituratus <- unicornfish 1279 observations
Acanthurus triostegus <- manini 466
Carcharhinus melanopterus <- black tip reef shark 51 obs
```{r}
species_list <- read_csv(here('Data','MCR_LTER_Fish_Species_List_20230608.csv'))
annual_survey <- read_csv(here('Data','MCR_LTER_Annual_Fish_Survey_20230615.csv'))

fishies <- annual_survey |> filter(Taxonomy == "Naso lituratus")
```



```{r}
#just the orange spine unicorn fishies
ggplot(fishies, aes(x = Year, y = Count)) +
  geom_line(color = "blue")
```
```{r}
#all fishies count
ggplot(annual_survey, aes(x = Year, y = Count)) +
  geom_line(color = "blue")
```
Link to data: https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-mcr.6.62
